NURM: a tool to locate numerical differences in pipelines

### 2019-10-09 paper.tex

Many experiments show that computational analysis results are still not 
completely reproducible. Computational environments including different 
operating systems, hardware categories, and software versions are known 
to have effects on the results produced by analysis pipelines. These 
effects are presumably due to the creation, propagation and 
amplification of small numerical differences across the pipelines. 
Although new techniques such as virtualization, version control 
systems, and provenance management tools provide more reliable 
environment and significantly improve the reproducibility of scientific 
findings. However, the precise causes of such instabilities and the 
path along which they propagate in the pipelines are unclear.  We 
present a technique to identify the processes in the pipeline that 
create numerical differences along the execution, and we apply this 
technique to the HCP structural pre-processing pipelines.

@@

A reproducible study provides a context in which one is able to get 
results that are consistent with the original 
work~\cite{plesser2018reproducibility}. Research findings are expected 
to be reproducible so that their authenticity and reliability can be 
evaluated.

Reproducibility is defined as the ability to regenerate the same 
results as the original findings when the experiment is reanalyzed by 
exactly the same analytic methods, software package, parameters, and 
data~\cite{peng2011reproducible}.

In addition, numerical reproducibility is defined as the ability to 
regenerate bit for bit identical results from multiple 
runs~\cite{hill2017numerical}. This Numerical reproducibility can be 
checked by comparing the binary content of the results which is 
computed by the checksum method.
Regarding these definitions, it must be pointed that a
computation might be reproducible, but not 
be numerically reproducible. For instance, small numerical differences 
created during the pipeline execution may hamper numerical 
reproducibility, but be negligible in the final results.

\note{Recently, scientists began to realize that the results of many 
scientific experiments were not reproducible.

In particular, validation of the reproducibility has been widely investigated 
in the field of neuroimaging in which uses optimized 
processing methods with the purpose of functional and/or structural 
assessments of the human brain.

According to the previous 
studies, variety of computational infrastructures including workstation 
types, parallelization methods, operating systems, and analysis 
packages are known to influence reproducibility~\cite{Gronenschild2012, 
diethelm2012limits, Glatard2015, bowring2019exploring}.

In particular, the effect of 
the operating system across some of the main 
neuroimaging pipelines including FSL, CIVET and FreeSurfer pipelines 
is studied in~\cite{Glatard2015, Gronenschild2012} that indicates a significant disparity. 
}

These irreproducibility issues are reported as the result of the 
creation, propagation, and amplification of small numerical 
differences~\cite{Gronenschild2012, diethelm2012limits, Glatard2015, 
bowring2019exploring}. In this case, the analysis pipelines are said to 
be numerically unstable. Numerical instability is a characteristic of 
the pipelines which amplify small numerical differences and then hamper 
the reproducibility of the analyses depending on the length of the 
pipeline and magnitude of the differences. In many cases, numerical 
instability is an important issue for reproducibility.

There are different solutions to make reproducible analysis including 
containerization techniques that encapsulate software/hardware 
dependencies, provenance capturing tools, and version control systems. 
However, a comprehensive solution requires to fix the numerical 
instabilities. For this purpose, we introduce a Numerical 
Reproducibility Measurement (NURM) tool to identify the processes in 
the pipeline that create numerical differences across several runs. In this 
paper, the execution results of pipelines across different operating 
systems are considered to find out the cause of the irreproducibility 
using interposition techniques.

@@

\tristan{define subjects, explain neuroimaging, bring context.}
\section{Tool description}

Figure~\ref{fig:overview} shows an overview of our method. For each tested
condition, Condition 1 and Condition 2, the pipeline is containerized with
Docker, and its interface is described with
Boutiques~\cite{glatard2017boutiques}. \reprozip~\cite{Chirigati2016} is
added to Condition 1 for provenance tracking. Our tool starts by
identifying temporary files and ``multi-write" files, i.e., files written
by more than one process (Figure~\ref{fig:overview-capturing}). To do that,
the tool first executes the pipeline in Condition 1, to obtain a process
graph (Figure~\ref{fig:overview-capturing} - (1)) and a list of result files.
The list of processes that create intermediary or multi-write
files is then generated (Figure~\ref{fig:overview-capturing} - (2)). 
Using this list, the tool modifies Docker image in Condition 1 and
exectutes the pipeline again through it, which turns to produce results
including temporary and overwritten files
(Figure~\ref{fig:overview-capturing} - (3,4)). In the second step, the
processes are labeled from the differences found in their outputs in the
two tested conditions (Figure~\ref{fig:overview-labelling}). To this order,
the tool builds docker image for Condition 2 by modifying pipeline
processes (Figure~\ref{fig:overview-labelling} - (1)) that recognised from 
the process graph. Then, the modified docker
image is executed in Condition 2 and each process is labeled right after
the running by comparing the results obtained in Condition 1
and output files of the process in Condition 2
(Figure~\ref{fig:overview-labelling} - (2)). We then label processes in 2
categories depending whether they
create differences (represented in red) or not (green). 
The tool replaces output file with the same file obtained in Condition 1 
if process creates differences (Figure~~\ref{fig:overview-labelling} - (3)).
Process labelling
is done through hooks which are defined during the docker image modifying
procedure, and incremental updates in Condition 2, until the end of the
pipeline execution.

@@

Different subjects may lead to different \emph{process tree}, due to
variations that may be coming from differences in data types or
cardinality. For instance, some images may have been acquired multiple
times in some subjects. Some of these differences can be neglected, for
instance when a data decompression step is present at the beginning of the
execution for some subjects only, and other ones cannot, when different
processing paths are used in different subjects.

Different \emph{process tree} shows that pipeline execution followed by a different procedure. 
This implies the possible impact of different subjects on pipeline results.
We cluster subjects into different types to specify how many different 
\emph{process trees} are generated. Clustering is performed based on the \emph{process trees} 
which nodes are labeled by the process name. 
In this order, 
the tree edit distance~\cite{zhang1989simple} is used as a similarity measure for 
representing the distance between labeled trees. Tree edit distance  is defined  
as the minimum number of edit operations which is needed to transform one tree into 
the other. Three edit operations are considered: node label modification, 
node removal, and node insertion. Each operation has an associated cost of 1.
With this distance, we cluster subjects using agglomerative hierarchical
clustering as implemented in SciPy~\cite{oliphant2007scipy}
(Algorithm~\ref{algo:hclustering}). As a result, the distance between
any two subjects in a cluster is lower than a threshold parameter.

Furthermore, we will show that even subjects in the same cluster, 
having the same topological \emph{process tree} structure, 
may have different labels as processes that create differences.

@@

We identified four types of subjects with different PreFreeSurfer process
trees (Table~\ref{table:data-clusters}). Differences between subject types
are coming from different  numbers of T1 and T2 images in the raw data:
subjects of type 1 have 2 T1 images and 2 T2 images, while subjects of type
4 only have 1 T1 image and 1 T2 image. \tristan{check that} This leads to
different sets of output files. We verified that the process trees were
identical for all subjects of the same type in different versions of
CentOS.

\begin{table}
\label{table:data-clusters}
\centering
\begin{threeparttable}
\caption{Clustering of 20 input subjects.}

\begin{tabular}{@{}llllll@{}}
\toprule
Type   & Number of Subjects  & Missing File(s)                                    \\ \midrule
type1  & 9                   & None                                               \\
type2  & 7                   & [``T1w/T1w2\_gdc.nii.gz",                          \\
       &                     &  ``T2w/T2w2\_gdc.nii.gz"]                          \\
type3  & 2                   & [``T1w/T1w2\_gdc.nii.gz"]                          \\
type4  & 2                   & [``T2w/T2w2\_gdc.nii.gz"]                          \\ \bottomrule
\end{tabular}
\begin{tablenotes}
     \small
     \item *Type refers to the topology type of the \emph{process tree} of subjects.
\end{tablenotes}
\end{threeparttable}
\end{table}

The frequency of the processes that create differences in 
the pipeline is shown in Figure~\ref{fig:pfs_freq}. 
\emph{FLIRT} is the most frequent process that produces differences.
Furthermore, the percentage of processes that introduce differences 
along with the 
sub-pipelines are depicted in this figure 
including linear registration with “\emph{FLIRT}” (in 
ACPC-Alignment, Brain Extraction, Distortion Correction, 
Atlas Registration), non-linear registration with “\emph{FNIRT}” (in 
Brain Extraction and Altas Registration), image warping with 
“\emph{new\_invwarp}” (in Brain Extraction and Atlas Registration). 
Besides, differences were observed in image mean and standard-deviation 
computations in image interpolation with “\emph{fslmaths}” (in Anatomical Average). 
Results show that \emph{fslmaths} and \emph{FLIRT} 
are the processes that at least have one subprocess to produce differences across all the subjects.

% \begin{figure}
%   \centering
%     \includegraphics[width=\columnwidth]{images/pfs_barchart.png} \caption{Number 
%     of occurrences of processes that create differences in PreFreeSurfer 
%     among 20 subjects.} 
%     \label{fig:pfs_chart}
%   \end{figure}

@@

We created checkerboard images from the image results of one subject in two OSes.
Alternating rectangular regions between two images, creating a characteristic 
checkerboard-like pattern of varying magnitudes.
Figure~\ref{fig:fnirt_result} shows differences in the files 
that are created as a part of ACPC-Alignment in PreFreeSurfer.
Each rectangle includes the pixels with different intensity values between images.

@@

In the first test, the results of PreFreeSurfer on CentOS 6
feed to the FreeSurfer pipeline on both OSes.
We fixed input data to determine whether the pipeline creates differences.
We processed 20 subjects using FreeSurfer.
NURM-tool identified \emph{mris\_make\_surfaces} as the only process in 
the pipeline that creates differences in 10 subjects.
This process generates surface files for cortical and white matter surfaces (see~\cite{} for more details).
Also, FreeSurfer did not introduce any differences in 10 subjects.

Results show that the static builds of FreeSurfer in which uses the embedded libraries, 
significantly avoids the creation of differences.
However, we found that between-OS differences occur when pre-processed data are different in FreeSurfer.

In the second test, the results of PreFreeSurfer feed to the FreeSurfer 
pipeline on each operating system separately.
Figure~\ref{fig:tissue_class} shows between-OS differences  
as a result of propagation and amplification of differences in FreeSurfer. % that coming from PreFreeSurfer outputs. 
This figure illustrates the sum of binarized differences 
from 20 subject results of the brain tissue segmentation across OSes.
Therefore, the magnitude of differences is in the range of 1-20 as the total number of subjects.

>>>>

Many experiments show that computational analysis results are still not 
completely reproducible. Computational environments including different 
operating systems, hardware categories, and software versions are known 
to have effects on the results produced by analysis pipelines. These 
effects are presumably due to the creation, propagation and 
amplification of small numerical differences across the pipelines. 
Although new techniques such as virtualization, version control 
systems, and provenance management tools provide a more reliable 
environment and significantly improve the reproducibility of scientific 
findings. However, the precise causes of such instabilities and the 
path along which they propagate in the pipelines are unclear.  We 
present a technique to identify the processes in the pipeline that 
create numerical differences along with the execution, and we apply this 
technique to the HCP structural pre-processing pipelines.

@@

Reproducibility is a crucial element of the scientific works, 
as it enables researchers to evaluate authenticity and reliability 
of the findings~\cite{plesser2018reproducibility}.
Reproducibility is defined as the ability to regenerate the same 
results as the original findings when the experiment is reanalyzed by 
the same analytic methods, software package, parameters, and 
data~\cite{peng2011reproducible}. 
In addition, numerical reproducibility is defined as the ability to 
regenerate bit for bit identical results from multiple 
runs~\cite{hill2017numerical}. 
We considered the numerical reproducibility in our experiments by comparing 
the binary content of the results using the checksum method.

Recently, validation of the reproducibility has been widely investigated 
in the field of neuroimaging in which uses optimized 
processing methods with the purpose of functional and structural 
assessments of the human brain.
According to the previous 
studies, the variety of computing infrastructures including workstation 
types, parallelization methods, operating systems, and analysis 
packages are known to influence the reproducibility of the analyses~\cite{Gronenschild2012, 
diethelm2012limits, Glatard2015, bowring2019exploring}.

In particular, the studies on the effect of the operating system on the results show that 
computational pipelines create small numerical differences~\cite{Glatard2015, Scaria2017}.
These differences mainly correspond to the mathematical functions implemented 
in different operating system libraries.
For instance, changing the mathematical functions like \emph{expf()} and 
\emph{cosf()} which manipulate the precision of floating-point representations, 
between \emph{glibc} libraries in different operating systems can produce 
small numerical differences.
Besides, a similar issue is expected for any operating system which is 
based on \emph{glibc}, the GNU C library.

These irreproducibility issues are reported as the result of the 
creation, propagation, and amplification of small numerical 
differences~\cite{Gronenschild2012, diethelm2012limits, Glatard2015, 
bowring2019exploring}. In this case, the analysis pipelines are said to 
be numerically unstable. Numerical instability is a characteristic of 
the pipelines which amplify small numerical differences and then hamper 
the reproducibility of the analyses depending on the length of the 
pipeline and magnitude of the differences. In many cases, numerical 
instability is an essential issue for reproducibility.

There are different solutions to improve the reproducibility of the analysis, 
including containerization techniques that encapsulate software/hardware 
dependencies, provenance capturing tools, and version control systems. 
However, a comprehensive solution requires to fix the numerical 
instabilities instead of masking the issue. For this purpose, we introduce 
the Numerical Reproducibility Measurement (NURM) tool to identify the processes in 
the pipeline that create numerical differences across several runs. 
Furthermore, we can fix the instability of the identified processes 
in the next steps.

@@

\section{Tool description}

Figure~\ref{fig:overview} shows an overview of our method. For each tested
condition, Condition 1 and Condition 2, the pipeline is containerized with
Docker, and its interface is described with
Boutiques~\cite{glatard2017boutiques}. \reprozip~\cite{chirigati2016reprozip} is
added to Condition 1 for provenance tracking. Our tool starts by
identifying temporary files and ``multi-write" files, i.e., files written
by more than one process (Figure~\ref{fig:overview-capturing}). To do that,
the tool first executes the pipeline in Condition 1, to obtain a process
graph (Figure~\ref{fig:overview-capturing} - (1)) and a list of result files.
The list of processes that create intermediary or multi-write
files is then generated (Figure~\ref{fig:overview-capturing} - (2)). 
Using this list, the tool modifies Docker image in Condition 1 and
exectutes the pipeline again through it, which turns to produce results
including temporary and overwritten files
(Figure~\ref{fig:overview-capturing} - (3,4)). In the second step, the
processes are labeled from the differences found in their outputs in the
two tested conditions (Figure~\ref{fig:overview-labelling}). To this order,
the tool builds docker image for Condition 2 by modifying pipeline
processes (Figure~\ref{fig:overview-labelling} - (1)) that recognised from 
the process graph. Then, the modified docker
image is executed in Condition 2 and each process is labeled right after
the running by comparing the results obtained in Condition 1
and output files of the process in Condition 2
(Figure~\ref{fig:overview-labelling} - (2)). We then label processes in 2
categories depending whether they
create differences (represented in red) or not (green). 
The tool replaces output file with the same file obtained in Condition 1 
if process creates differences (Figure~\ref{fig:overview-labelling} - (3)).
Process labelling
is done through hooks which are defined during the docker image modifying
procedure, and incremental updates in Condition 2, until the end of the
pipeline execution.

@@

Different subjects may lead to different process tree, due to
variations that may be coming from differences in data types or
cardinality. For instance, some images may have been acquired multiple
times in some subjects. Some of these differences can be neglected, for
instance when a data decompression step is present at the beginning of the
execution for some subjects only, and other ones cannot, when different
processing paths are used in different subjects.

Different process tree shows that pipeline execution followed by a different procedure. 
This implies the possible impact of different subjects on pipeline results.
We cluster subjects into different types to specify how many different 
process trees are generated. Clustering is performed based on the process trees 
which nodes are labeled by the process name. 
In this order, 
the tree edit distance~\cite{zhang1989simple} is used as a similarity measure for 
representing the distance between labeled trees. Tree edit distance  is defined  
as the minimum number of edit operations which is needed to transform one tree into 
the other. Three edit operations are considered: node label modification, 
node removal, and node insertion. Each operation has an associated cost of 1.
With this distance, we cluster subjects using agglomerative hierarchical
clustering as implemented in SciPy~\cite{oliphant2007scipy}
(Algorithm~\ref{algo:hclustering}). As a result, the distance between
any two subjects in a cluster is lower than a threshold parameter.

Furthermore, we will show that even subjects in the same cluster, 
having the same topological process tree structure, 
may have different labels as processes that create differences.

@@

We identified four types of subjects with different PreFreeSurfer process
trees (Table~\ref{table:data-clusters}). Differences between subject types
are coming from different  numbers of T1 and T2 images in the raw data:
subjects of type 1 have 2 T1 images and 2 T2 images, while subjects of type
2 only have 1 T1 image and 1 T2 image. This leads to
different sets of output files. We verified that the process trees were
identical for all subjects of the same type in different versions of
CentOS.

\begin{table}
\centering
\begin{threeparttable}
\caption{Clustering of 20 input subjects.}
\label{table:data-clusters}

\begin{tabular}{@{}llllll@{}}
\toprule
Type   & Number of Subjects  & Missing File(s)                                    \\ \midrule
type1  & 9                   & None                                               \\
type2  & 7                   & [``T1w", ``T2w"]                                   \\
type3  & 2                   & [``T1w"]                                           \\
type4  & 2                   & [``T2w"]                                           \\ \bottomrule
\end{tabular}
\begin{tablenotes}
     \small
     \item *Type refers to the topology type of the process tree of subjects.
\end{tablenotes}
\end{threeparttable}
\end{table}

Figure~\ref{fig:pfs_freq} shows the frequency of the processes that create differences 
in the pipeline (the number in the bracket).
Furthermore, this figure represents the percentage of the processes that introduce 
differences along with the sub-pipelines, including linear registration 
with “\emph{FLIRT}” (in ACPC-Alignment, Brain Extraction, Distortion Correction, 
Atlas Registration), non-linear registration with “\emph{FNIRT}” (in Brain Extraction 
and Altas Registration), image warping with “\emph{new\_invwarp}” (in Brain Extraction 
and Atlas Registration). 
Besides, differences were observed in image mean and standard-deviation 
computations in image interpolation with “\emph{fslmaths}” (in Anatomical Average).

Results show that \emph{FLIRT} is the most frequent process that produces 
differences in ACPC-Alignment across all the subjects, Distortion Correction 
in 95\% of the subjects, and Atlas Registration in 85\% of the subjects. 
In addition, there is a \emph{fslmaths} process in Anatomical Average that 
creates differences in all 11 subjects in which it has appeared.

@@

To visually represent the magnitude of variations, we created checkerboard 
images from the image results of one subject in two OSes. Figure~\ref{fig:fnirt_result} 
shows the checkerboard images from the results of ACPC-Alignment in PreFreeSurfer. 
Each rectangle includes the replaced pixels with different intensity values between images.

%Alternating rectangular regions between two images, creating a characteristic 
%checkerboard-like pattern of varying magnitudes.

@@

In the first test, the results of PreFreeSurfer on CentOS 6
feed to the FreeSurfer pipeline on both OSes. 
We fixed input data to determine whether the pipeline creates differences.
We processed 20 subjects using FreeSurfer.
NURM-tool identified \emph{mris\_make\_surfaces} as the only process in 
the pipeline that creates differences among 10 subjects.
This process generates surface files for cortical and white matter surfaces 
(see~\cite{fischl2012freesurfer} for more detail of the process). 
Also, FreeSurfer did not introduce any differences in 10 subjects.

Results show that the static builds of FreeSurfer in which uses the embedded libraries, 
significantly avoids the creation of differences.
However, we found that between-OS differences occur when pre-processed data are different in FreeSurfer. 
For this purpose, in the second test, the results of PreFreeSurfer feed to the FreeSurfer 
pipeline on each operating system separately.
Figure~\ref{fig:tissue_class} shows between-OS differences  
as a result of propagation and amplification of differences in FreeSurfer. % that coming from PreFreeSurfer outputs. 
This figure illustrates the sum of binarized differences 
from 20 subject results of the brain tissue segmentation across OSes.
The magnitude of differences is in the range of 1-20 as the total number of subjects.
